---
title: "SHAR"
author: "Maximilian H.K. Hesselbarth"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# SHAR

**S**pecies **H**abitat **A**ssociations in **R** is a package to analyse species habitat associations using the spatial point pattern of the species and a raster layer with habitats. All techniques are based on Plotkin et al. (2000), Harms et al. (2001) and Hesselbarth et al. (in prep.). 

## Methods
The null model is that there are no dependencies between the point pattern and the habitats. To test for this, a randomization breaks possible dependencies between the two, while keeping the spatial structure. 

Three different methods are availaible: 

* $\Gamma$-Test
* Torus-translation tests
* Pattern reconstruction

## Example

```{r, echo = FALSE, message = FALSE}
  library(raster)
  library(spatstat)
```
  
First, we use a dataset of the `spatstat` packages.  
```{r}  
  pattern <- bei
  habitats <- bei.extra$elev
````

### References

Plotkin, J.B., Potts, M.D., Leslie, N., Manokaran, N., LaFrankie, J.V., Ashton, P.S., 2000. Species-area curves, spatial aggregation, and habitat specialization in tropical forests. J. Theor. Biol. 207, 81–99.

Harms, K.E., Condit, R., Hubbell, S.P., Foster, R.B., 2001. Habitat associations of trees and shrubs in a 50-ha neotropical forest plot. J. Ecol. 89, 947–959.



