---
title: "Analysing the climatic niche of Cormus domestica"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analysing the climatic niche of Cormus domestica}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}

# knitr::opts_chunk$set(out.width='750px', dpi=200)

```

## Outline

This vignette demonstrates how to use `shar` to analyse species occurrence data obtained from the Global Biodiversity Information Facility (GBIF) and environmental raster data obtained from the Climate Research Unit (CRU) entirely in `R`, using the "pattern reconstruction" approach as detailed in the background vignette. The distribution of the tree species *Cormus domestica* in Europe is selected, a tree which tolerates a wide range of conditions but favors warm to mild climates, and is most commonly found in the FAO ecological zone "Subtropical Dry Forest", which occupies much of the Mediterranean coast (De Rigo et al., 2016).

## Load required packages

```{r packages, echo=TRUE, message=FALSE, warning=FALSE}

library(dplyr) # For data wrangling
library(rgbif) # For retrieving species occurrence data 
library(rnaturalearth) # For retrieving geographical data
library(getCRUCLdata) # For retrieving climate raster data
library(sf) # For spatial data operations
library(raster) # For spatial data operations
library(terra) # For spatial data operations
library(shar) # For species-habitat association analysis
library(spatstat) # For spatial point pattern analysis
library(spatstat.geom) # For spatial point pattern analysis
library(patchwork) # For composing multiple plots

```

## Download occurrence data

To retrieve species occurrence data the `R` package `rgbif` (Chamberlain & Boettiger, 2017) is used, which provides an interface to access the GBIF database.

```{r gbif, echo=TRUE, message=FALSE, warning=FALSE}

# Retrieve key for Cormus domestica
key <- rgbif::name_backbone(name = 'Cormus domestica', kingdom = 'plants')

# Retrieve occurrences
res <- rgbif::occ_search(taxonKey = as.numeric(key$usageKey), limit = 99999)

# Create a simple data frame containing only the unique identifier (key),
# latitude (lat), and longtitude (lon).
data_simp <- data.frame(key = res$data$key, 
                        lat = res$data$decimalLatitude, 
                        lon = res$data$decimalLongitude) |> 
  dplyr::filter(!is.na(lat) | !is.na(lon))

```

## Download map data

Spatial polygon data for the world is obtained from the `rnaturalearth` package (South, 2022), the map is then restricted to the European region. The `spatstat` package requires geospatial data in the format of a projected coordinate system; the data is therefore converted from the geographic coordinate system [4336](https://epsg.io/4326) to the projected coordinate system [3395](https://epsg.io/3395). The `shar` function `reconstruct_pattern` requires a spatial point pattern (ppp) object bounded within an observation window of the class `owin`, which is then created.

```{r maps, echo=TRUE, message=FALSE, warning=FALSE}

# Retrieve data from rnaturalearth
# A scale of 50 is used, as it provides sufficient detail, without later slowing 
# down shar::reconstruct pattern too much.
worldmap <- rnaturalearth::ne_countries(returnclass = "sf", scale = 50) |>
  sf::st_transform(crs = sf::st_crs(3395))

# Manually establish bounding box
eur_box <- sf::st_sfc(sf::st_point(c(-20, 30)),
                      sf::st_point(c(45, 73)),
                      crs = 4326) |>
  sf::st_transform(crs = sf::st_crs(3395))

# Crop world map to include polygons within Europe extent
eur <- sf::st_crop(x = worldmap, y = eur_box)

# Define observation window
eur_owin <- spatstat.geom::as.owin(eur$geometry)
```

## Download climate data

The environmental variable selected for demonstrative purposes is the mean temperature in June over the 1961-1990 period. Data is obtained through the `getCRUCLdata` package (Sparks, 2017) which provides access to the datasets described in (New et al., 2002).

```{r cru_data, echo=TRUE, message=FALSE, warning=FALSE}

# Download data as a raster brick through the getCRUCLdata package
# Mean temperature (tmn) data should be 180.4MB
cru_data <- getCRUCLdata::get_CRU_stack(pre = FALSE,
                                        pre_cv = FALSE,
                                        rd0 = FALSE,
                                        tmp = TRUE,
                                        dtr = FALSE,
                                        reh = FALSE,
                                        tmn = FALSE,
                                        tmx = FALSE,
                                        sunp = FALSE,
                                        frs = FALSE,
                                        wnd = FALSE,
                                        elv = FALSE,
                                        cache = FALSE)


```

## Prepare landscape raster

The climate data obtained above is restricted to the European region, then classified into 10 habitats, as set in the `fixedBreaks` argument of the `classify_habitats` function.

```{r landscape_ras, echo=TRUE, message=FALSE, warning=FALSE}

# Select month
tmp_raster_jun <- cru_data$tmp$jun

# Crop tmp raster
tmp_raster_jun_eur <- terra::crop(x = tmp_raster_jun, 
                                  y = c(xmin=-20, xmax=45, ymin=30, ymax=73))

# Reproject raster
tmp_raster_jun_eur_3395 <- raster::projectRaster(tmp_raster_jun_eur, crs = 3395)

# Classify landscape
landscape_classified <- shar::classify_habitats(raster = tmp_raster_jun_eur_3395,
                                                return_breaks = TRUE,
                                                style = "fixed",
                                                fixedBreaks = c(0, 5, 7.5,
                                                                10, 12.5, 15, 
                                                                17.5, 20, 25, 
                                                                30, 35))

```

```{r land_plots, echo=FALSE, message=FALSE, warning=FALSE, fig.retina=FALSE, out.width="100%", dpi=400}

raster_unclassed_df <- terra::as.data.frame(tmp_raster_jun_eur_3395, xy = TRUE) |>
  dplyr::rename("value" = "jun") |> 
  dplyr::mutate("type" = "Unclassified")

raster_classed_df <- terra::as.data.frame(landscape_classified$raster, xy = TRUE) |>
  dplyr::rename("value" = "layer") |>
  dplyr::mutate("type" = "Classified")

plot_unclassed <- ggplot2::ggplot() +
  ggplot2::geom_raster(data = raster_unclassed_df,
                       mapping = ggplot2::aes(x = x, y = y, fill = value)) +
  ggplot2::geom_sf(data = eur$geometry,
                   mapping = ggplot2::aes(),
                   colour = "black", fill = NA, size = 0.1) +
  ggplot2::facet_wrap(~type) + 
  ggplot2::theme_minimal() +
  ggplot2::xlab(label = NULL) +
  ggplot2::ylab(label = NULL) +
  ggplot2::labs(fill = NULL) +
  ggplot2::scale_fill_distiller(palette = "RdBu", 
                                na.value = "transparent") +
  ggplot2::theme(panel.grid.major = ggplot2::element_line(colour = "#c9c9c9", 
                                                          linetype = "dashed", 
                                                          size = 0.075), 
                 panel.background = ggplot2::element_rect(fill = "#f0f8ff"), 
                 panel.border = ggplot2::element_rect(fill = NA),
                 text = ggplot2::element_text(size = 6),
                 axis.text.x = ggplot2::element_text(size = 4),
                 axis.text.y = ggplot2::element_text(size = 4),
                 plot.margin = ggplot2::margin(t = 0,  # Top margin
                                               r = 0,  # Right margin
                                               b = 0,  # Bottom margin
                                               l = 0),
                 # legend.position = "bottom",
                 legend.position = c(0.5, -0.2), 
                 legend.direction = "horizontal",
                 legend.justification = "center",
                 legend.text = ggplot2::element_text(size = 3),
                 legend.key.size = ggplot2::unit(0.125, 'cm'),
                 legend.key.width = ggplot2::unit(0.3, "cm")
                 ) +
  NULL

plot_classed <- ggplot2::ggplot() +
  ggplot2::geom_raster(data = raster_classed_df,
                       mapping = ggplot2::aes(x = x, y = y,
                                              fill = factor(value)
                                              # fill = value
                                              )) +
  ggplot2::geom_sf(data = eur$geometry,
                   mapping = ggplot2::aes(),
                   colour = "black", fill = NA, size = 0.1) +
  ggplot2::facet_wrap(~type) + 
  ggplot2::theme_minimal() +
  ggplot2::xlab(label = NULL) +
  ggplot2::ylab(label = NULL) +
  ggplot2::labs(fill = NULL) +
  ggplot2::scale_fill_brewer(palette = "RdBu", 
                             direction = -1,
                             na.value = "transparent", 
                             guide = ggplot2::guide_legend()) +
  ggplot2::theme(panel.grid.major = ggplot2::element_line(colour = "#c9c9c9", 
                                                          linetype = "dashed", 
                                                          size = 0.075), 
                 panel.background = ggplot2::element_rect(fill = "#f0f8ff"), 
                 panel.border = ggplot2::element_rect(fill = NA),
                 text = ggplot2::element_text(size = 6),
                 axis.text.x = ggplot2::element_text(size = 4),
                 axis.text.y = ggplot2::element_blank(), # ggplot2::element_text(size = 4),
                 plot.margin = ggplot2::margin(t = 0,  # Top margin
                                               r = 0,  # Right margin
                                               b = 0,  # Bottom margin
                                               l = 0.25),
                 # legend.position = "bottom",
                 legend.position = c(0.5, -0.2),
                 legend.direction = "horizontal",
                 legend.justification = "center",
                 legend.text = ggplot2::element_text(size = 3),
                 legend.key.size = ggplot2::unit(0.125, 'cm'),
                 legend.key.width = ggplot2::unit(0.3, "cm")
                 ) +
  NULL

plot_unclassed + plot_classed

```

## Prepare occurrence data

The occurrence data is prepared, then the `shar` function `reconstruct_pattern` is called, yielding the randomised occurrence data within the observation window as required by the `results_habitat_association` function.

```{r occ_prep, echo=TRUE, message=FALSE, warning=FALSE}

# Convert occurrence data to a simple features object
data_sf <- sf::st_as_sf(data_simp, 
                        coords = c("lon", "lat"),
                        crs = 4326)

# Restrict occurences to those within the European region, then re-project the data
data_sf_eur_3395 <- sf::st_crop(x = data_sf,
                                y = c(xmin=-20, xmax=45, ymin=30, ymax=73)) |>
  sf::st_transform(crs = sf::st_crs(3395))

# Extract the coordinates as a matrix from the sf occurrences object
data_sf_eur_coords <- sf::st_coordinates(data_sf_eur_3395)

# Create a spatiral points pattern object containing the occurrence data
data_sf_eur_ppp <- spatstat.geom::as.ppp(X = data_sf_eur_coords, 
                                         W = eur_owin)

# Reconstruct pattern
recon_pattern <- shar::reconstruct_pattern(pattern = data_sf_eur_ppp,
                                           max_runs = 1)

```

```{r occ_plots, echo=FALSE, message=FALSE, warning=FALSE, fig.retina=FALSE, out.width="100%", dpi=400}

recon_occ_df <- as.data.frame(recon_pattern$randomized$randomized_1) |> 
  dplyr::mutate("type" = "Randomised Occurrences")

real_occ_df <- data_sf_eur_coords |>
  as.data.frame() |> 
  dplyr::rename(x = "X",
                y = "Y") |> 
  dplyr::mutate("type" = "Real Occurrences")

real_occ_plot <- ggplot2::ggplot() +
  ggplot2::geom_sf(data = eur$geometry,
                   mapping = ggplot2::aes(),
                   colour = "black", fill = "white", size = 0.01) +
  ggplot2::geom_point(real_occ_df, 
                      mapping = ggplot2::aes(x = x, y = y),
                      size = 0.2, stroke = 0, shape = 16, color = "Red") +
  ggplot2::facet_wrap(~type) + 
  ggplot2::theme_minimal() +
  ggplot2::xlab(label = NULL) +
  ggplot2::ylab(label = NULL) +
  ggplot2::theme(panel.grid.major = ggplot2::element_line(colour = "#c9c9c9", 
                                                          linetype = "dashed", 
                                                          size = 0.075), 
                 panel.background = ggplot2::element_rect(fill = "#f0f8ff"), 
                 panel.border = ggplot2::element_rect(fill = NA),
                 text = ggplot2::element_text(size = 6),
                 axis.text.x = ggplot2::element_text(size = 4),
                 axis.text.y = ggplot2::element_text(size = 4),
                 plot.margin = ggplot2::margin(t = 0,  
                                               r = 0,  
                                               b = 0,  
                                               l = 0)) + 
  NULL

recon_occ_plot <- ggplot2::ggplot() +
  ggplot2::geom_sf(data = eur$geometry,
                   mapping = ggplot2::aes(),
                   colour = "black", fill = "white", size = 0.01) +
  ggplot2::geom_point(recon_occ_df, 
                      mapping = ggplot2::aes(x = x, y = y),
                      size = 0.2, stroke = 0, shape = 16, color = "Red") +
  ggplot2::facet_wrap(~type) + 
  ggplot2::theme_minimal() +
  ggplot2::xlab(label = NULL) +
  ggplot2::ylab(label = NULL) +
  ggplot2::theme(panel.grid.major = ggplot2::element_line(colour = "#c9c9c9", 
                                                          linetype = "dashed", 
                                                          size = 0.075), 
                 panel.background = ggplot2::element_rect(fill = "#f0f8ff"), 
                 panel.border = ggplot2::element_rect(fill = NA),
                 text = ggplot2::element_text(size = 6),
                 axis.text.x = ggplot2::element_text(size = 4),
                 axis.text.y = ggplot2::element_blank(), # ggplot2::element_text(size = 4),
                 plot.margin = ggplot2::margin(t = 0,  
                                               r = 0,  
                                               b = 0,  
                                               l = 0)) + 
  NULL

real_occ_plot + recon_occ_plot

```


## Results

The analysis function `results_habitat_association` is then called. The results of the pattern reconstruction analysis show that *Cormus domestica* is positively associated with locations which experience a mean June temperature of 15C-17.5C & 17.5C-20C.

```{r model_run, echo=TRUE, message=FALSE, warning=FALSE}

# Establish significance level
sig_level <- 0.01

# Run analysis
results <- shar::results_habitat_association(pattern = recon_pattern, 
                                             raster = landscape_classified$raster,
                                             breaks = landscape_classified$breaks,
                                             significance_level = sig_level) |> 
  dplyr::arrange(habitat)

results

```

## References

Chamberlain SA, Boettiger C. 2017. R Python, and Ruby clients for GBIF species occurrence data. PeerJ Preprints 5:e3304v1 <doi:10.7287/peerj.preprints.3304v1>

De Rigo, D., Caudullo, G., Houston Durrant, T. and San-Miguel-Ayanz, J., 2016. The European Atlas of Forest Tree Species: modelling, data and information on forest tree species. *European Atlas of Forest Tree Species*, p.e01aa69. <doi:10.2788/4251>

New, M., Lister, D., Hulme, M. and Makin, I., 2002. A high-resolution data set of surface climate over global land areas. *Climate research*, *21*(1), pp.1-25. <doi:10.3354/cr021001>

South A (2022). *rnaturalearth: World Map Data from Natural Earth*. <https://docs.ropensci.org/rnaturalearth> (website) <https://github.com/ropensci/rnaturalearth.>

Sparks, (2017). getCRUCLdata: Use and Explore CRU CL v. 2.0 Climatology Elements in R. Journal of Open Source Software, 2(12), 230, <doi:10.21105/joss.00230>
