---
title: "Parallelization"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parallelization}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The `shar` packages has no build-in parallelization, however, there are many `R` frameworks that allow to run code in parallel or on high-performance clusters (see e.g., [`future`]( https://CRAN.R-project.org/package=future), [`clustermq`]( https://CRAN.R-project.org/package=clustermq) or [`rslurm`]( https://CRAN.R-project.org/package=rslurm)). Thus, `shar` provides utility functions that facilitate its usage together with any parallelization framework. 

The following examples illustrates how to use `future` package to randomize patterns using `fit_point_process` in parallel using all available cores on a local machine. Similarly, the core idea of the following code could be used to run `shar` on a high performance cluster.

First, we need to load all required packages. This includes `future` and `future.apply`.

```{r load-packages, message = FALSE}
library(shar)

library(future)
library(future.apply)
```

The `future` packages allows to run code in parallel using only a few lines of code. By setting the `future` plan to `multisession`, the package automatically resolves all following `futures` in parallel.

Importantly with this approach, you need one randomization per core (`n_random = 1`) and set `simplify = TRUE` to return the point pattern only. This results in a list of randomized point patterns.

```{r parallel}
future::plan(multisession)

fitted_list <- future.apply::future_lapply(X = 1:39, FUN = function(i) {
   shar::fit_point_process(pattern = shar::species_b, n_random = 1,
                          return_input = FALSE, simplify = TRUE, verbose = FALSE)
}, future.seed = 42)
```

Next, you can use the `list_to_randomized()` function to convert this list of randomized pattern to a `rd_pat` object that will work will all other functions of the `shar` package.

```{r convert-list}
fitted_rd <- shar::list_to_randomized(list = fitted_list, observed = shar::species_b)
```

Lastly, the created objects can be used to analyse if species-habitat associations are present as usual.

```{r results}
landscape_classified <- shar::classify_habitats(raster = landscape)

results_habitat_association(pattern = fitted_rd, raster = landscape_classified)
```

Of course, this idea can be used to randomize the raster data as well. Furthermore, any other parallelization framework could be used.
