---
title: "How to reconstruct multiple patterns"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to reconstruct multiple patterns}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

In case you want to reconstruct several patterns at once (e.g. for different points in time if repeated censuses are available), you can use the following code.

```{r load-packages, message = FALSE}
# library(tidyverse)
library(shar)
library(spatstat)
```

In case you want to only create the spatial characteristics, this is straightforward using `lapply()`. 

```{r several-patterns}
# create list with patterns
list_pattern <- list(species_a, species_b)

# reconstruct all patterns in list
result <- lapply(list_pattern, function(x) reconstruct_pattern(pattern = x, 
                                                               n_random = 3, 
                                                               verbose = FALSE))

```

The result will be a nested list including all *m* randomization (including the observed pattern) of the *n* provided input patterns. 

```{r result-spatial}
# get mean energy
lapply(result, function(x) calculate_energy(pattern = x,
                                            verbose = FALSE))
```

Another possible would be to first reconstruct *n* times the spatial characteristics and afterwards reconstruct the marks *m* times for each of the *n* spatial reconstructions.

Firstly, reconstruct only the spatial characteristics *n* times. The observed pattern is not needed in this case, so you can put `return_input = FALSE`.

```{r reconstruct-pattern}
# reconstruct spatial strucutre
reconstructed_pattern <- reconstruct_pattern(species_a, 
                                             n_random = 3, 
                                             return_input = FALSE,
                                             verbose = FALSE)
```

Secondly, to reconstruct the (numeric) marks of the observed pattern for each of the spatially reconstructed patterns, just use `lapply()` in combination with `reconstruct_marks()`.

```{r reconstruct-marks}
# get only selected marks of input (numeric marks)
species_a_marks <- spatstat::subset.ppp(species_a, select = dbh)

# reconstruct marks 3 times for each input pattern
result_marks <- lapply(reconstructed_pattern, 
                       function(x) reconstruct_marks(pattern = x, 
                                                     marked_pattern = species_a_marks, 
                                                     n_random = 3, verbose = FALSE))
```

Again, the result is a nested list with the same dimensions as provided input patterns and reconstructions.

```{r result-marks}
# get energy
lapply(result_marks, function(x) calculate_energy(pattern = x, 
                                                  verbose = FALSE))
```
